Legion: Expressing Locality and Independence with Logical Regions

### 2012-06-30 doc/popl2013/langdef.tex
more stuff (not all of it good)

@@

>>>>

\finfrule
{\begin{array}{l}
E_1 \sim H \hspace{1cm}
E_2 \sim H
\end{array}}
{ E_1 \dplus E_2 \sim H } \\

@@

\subsection{Noninterference}

For a given expression $e$, we can consider $v^o$ to be a ``sequential'' result and $E^o$ to be the corresponding ``sequential'' effects if: \\

\framebox{$M, L, H, S, \emptyset \vdash e \mapsto v^o, E^o$} \\

(i.e. if there are no manipulations of the store other than what's done by
the evaluation of the expression $e$.)  We now wish to consider what parallel
evaluations of this expression with other expression evaluations will still
yield our ``sequential'' result. \\

Two effects are considered to be noninterfering, written $E_1 \# E_2$ if for
every $\epsilon_1$ in $E_1$ and every $\epsilon_2$ in $E_2$,
\begin{enumerate}
\item $\epsilon_1$ and $\epsilon_2$ are to different memory locations, or
\item $\epsilon_1 = read(l, v_1)$ and $\epsilon_2 = read(l, v_2)$, or
\item $\epsilon_1 = reduce_{id}(l, v_1)$ and $\epsilon_2 = reduce_{id}(l, v_2)$
\end{enumerate}

We define nonintereference (under a given mapping $M$) of static privileges
$\Phi_1$ and $\Phi_2$ similarly.  We use the following table for determining
noninterference of two individual priviledges $\phi_1$ and $\phi_2$:

\begin{tabular}{ccccc}
\multicolumn{2}{l}{$\phi_1 \#_M \phi_2 ?$} & \multicolumn{3}{c}{$\phi_2$} \\
\multicolumn{2}{l}{} & $reads(r_2)$ & $writes(r_2)$ & $reduces(id_2, r_2)$ \\
\multirow{3}{*}{$\phi_1$} & $reads(r_1)$ & true & $M(r_1) \cap M(r_2) = \emptyset$ & $M(r_1) \cap M(r_2) = \emptyset$\\
& $writes(r_1)$ & $M(r_1) \cap M(r_2) = \emptyset$ & $M(r_1) \cap M(r_2) = \emptyset$ & $M(r_1) \cap M(r_2) = \emptyset$ \\
& $reduces(id_1, r_1)$ & $M(r_1) \cap M(r_2) = \emptyset$ & $M(r_1) \cap M(r_2) = \emptyset$ & $\begin{array}{l}id_1 = id_2 \vee \\ M(r_1) \cap M(r_2) = \emptyset\end{array}$
\end{tabular}

We can then define: \\

\framebox{$\Phi_1 \#_M \Phi_2 \Leftrightarrow \displaystyle\bigwedge_{\begin{array}{c}\phi_1 \in \Phi_1 \\ \phi_2 \in \Phi_2 \end{array}} \phi_1 \#_M \phi_2$} \\

From the soundness of effects, result, we can show that for two expressions
$e_1$ and $e_2$, if their privileges are noninterfering, any potential memory
traces from their evaluation must also be noninterfering.  Formally:

\finfrule
{\begin{array}{l}
\Gamma, \Phi_1, \Omega \vdash e_1 : T_1 \\
\Gamma, \Phi_2, \Omega \vdash e_2 : T_2 \\
\opsenv e_1 \mapsto v_1, E_1 \\
\opsenv e_2 \mapsto v_2, E_2 \\
\Phi_1 \#_M \Phi_2
\end{array}}
{E_1 \# E_2}

(TODO: Generalize this further and allow $M_1 \not= M_2$?) \\

Note that this result holds even if $C \not= \emptyset$.  This allows 
two ``locally independent'' subtasks to run in parallel even if one or both
of them is interacting (in a programmer-permitted way) with a third subtask.

Noninterference of effects is something that can only be determined after
evaluation of an expression is completed, and only at great expense (see STM).
In contrast, the noninterference of privileges can be efficiently determined at
runtime.  Even though the privileges themselves are static, the
mapping $M$ can be dynamically determined, allowing noninterference to be 
judged in many more cases than a purely static analysis can achieve.

(TODO: Talk about coherence modes and how the runtime orders subtasks.  Maybe
describe it in terms of a ``canonical'' memory trace and then allow
reorderings?)

\subsection{Hierarchical Scheduling}

Let $e_1$ and $e_2$ be expressions that require privileges $\Phi_1$ and
$\Phi_2$.  Now let $e_3$ be some subexpression of $e_1$ and $e_4$ be some 
subexpression of $e_2$, with corresponding privileges $\Phi_3$ and $\Phi_4$.

Proper scheduling of the evaluation of $e_3$ and $e_4$ depends on whether
they interfere.  If they do intefere (and must therefore be serialized), we
will have $E_3 \cancel\# E_4$, and by the above result, $\Phi_3 \cancel\# \Phi_4$.

We now show that this will also imply that $\Phi_1 \cancel\# \Phi_2$, and
unless sequential behavior has been relaxed by the programmer, the
evaluations of $e_1$ and $e_2$ will be serialized.  Since the evaluation of
$e_3$ is a subsequence of the evaluation of $e_1$ and the evaluation of $e_4$
is a subsequence of the evaluation of $e_2$, the evaluations of $e_3$ and $e_4$
are properly serialized as well.

Conversely, if $e_1$ and $e_2$ are judged to be noninterfering, the schedule
does not need to check for interference between $e_3$ and $e_4$, as their
noninterference is guaranteed.

Proof is straightforward once armed with the following lemmas:

\framebox{$\Phi_1 \#_M \Phi_2 \Leftrightarrow \Phi_1^* \#_M \Phi_2^*$}

\framebox{$\Phi_1 \#_M \Phi_2 \wedge \Phi_3 \subseteq \Phi_2 \Rightarrow \Phi_1 \#_M \Phi_3$}
